Efficiently Handling Guards in the low level design of RPython's tracing JIT

### 2012-07-19 talk/vmil2012/paper.tex
- write about high-level handling of resume data - add a start of a bibliography

@@

- traces follow the execution path during tracing, other path not compiled at first
- points of possible divergence from that path are guards
- since path can later diverge, at the guards it must be possible to re-build interpreter state in the form of interpreter stack frames
- tracing does inlining, therefore a guard must contain information to build a whole stack of frames
- optimization rewrites traces, including removal of guards

- frames consist of a PC and local variables
- rebuild frame by taking local SSA variables in the trace and mapping them to variables in the frame

two forces:
- there are lots of guards, therefore the information must be stored in a compact way in the end
- tracing must be fast

@@

>>>>

Since tracing linearizes control flow by following one concrete execution,
not the full control flow of a program is observed.
The possible points of deviation from the trace are guard operations
that check whether the same assumptions as during tracing still hold.
In later executions of the trace the guards can fail.
If that happens, execution needs to continue in the interpreter.
This means it is necessary to attach enough information to a guard
to construct the interpreter state when that guard fails.
This information is called the \emph{resume data}.

To do this reconstruction, it is necessary to take the values
of the SSA variables of the trace
and build interpreter stack frames.
Tracing aggressively inlines functions.
Therefore the reconstructed state of the interpreter
can consist of several interpreter frames.

If a guard fails often enough, a trace is started from it
to create a trace tree.
When that happens another use case of resume data
is to construct the tracer state.

There are several forces guiding the design of resume data handling.
Guards are a very common operations in the traces.
However, a large percentage of all operations
are optimized away before code generation.
Since there are a lot of guards
the resume data needs to be stored in a very compact way.
On the other hand, tracing should be as fast as possible,
so the construction of resume data must not take too much time.

\subsection{Capturing of Resume Data During Tracing}
\label{sub:capturing}

Every time a guard is recorded during tracing
the tracer attaches preliminary resume data to it.
The data is preliminary in that it is not particularly compact yet.
The preliminary resume data takes the form of a stack of symbolic frames.
The stack contains only those interpreter frames seen by the tracer.
The frames are symbolic in that the local variables in the frames
do not contain values.
Instead, every local variables contains the SSA variable of the trace
where the value would later come from, or a constant.

\subsection{Compression of Resume Data}
\label{sub:compression}

The core idea of storing resume data as compactly as possible
is to share parts of the data structure between subsequent guards.
This is often useful because the density of guards in traces is so high,
that quite often not much changes between them.
Since resume data is a linked list of symbolic frames
often only the information in the top frame changes from one guard to the next.
The other frames can often be just reused.
The reason for this is that during tracing only the variables
of the currently executing frames can change.
Therefore if two guards are generated from code in the same function
the resume data of the rest of the stack can be reused.

\subsection{Interaction With Optimization}
\label{sub:optimization}

Guards interact with optimizations in various ways.
Most importantly optimizations try to remove as many operations
and therefore guards as possible.
This is done with many classical compiler optimizations.
In particular guards can be remove by subexpression elimination.
If the same guard is encountered a second time in the trace,
the second one can be removed.
This also works if a later guard is weaker implied by a earlier guard.

One of the techniques in the optimizer specific to tracing for removing guards
is guard strengthening~\cite{bebenita_spur:_2010}.
The idea of guard strengthening is that if a later guard is stronger
than an earlier guard it makes sense to move the stronger guard
to the point of the earlier, weaker guard and to remove the weaker guard.
Moving a guard to an earlier point is always valid,
it just means that the guard fails earlier during the trace execution
(the other direction is clearly not valid).

The other important point of interaction between resume data and the optimizer
is RPython's allocation removal optimization~\cite{bolz_allocation_2011}.
This optimization discovers allocations in the trace that create objects
that do not survive long.
An example is the instance of \lstinline{Even} in the example\cfbolz{reference figure}.
Allocation removal makes resume data more complex.
Since allocations are removed from the trace it becomes necessary
to reconstruct the objects that were not allocated so far when a guard fails.
Therefore the resume data needs to store enough information
to make this reconstruction possible.

Adding this additional information is done as follows.
So far, every variable in the symbolic frames
contains a constant or an SSA variable.
After allocation removal the variables in the symbolic frames can also contain
``virtual'' objects.
These are objects that were not allocated so far,
because the optimizer removed their allocation.
The virtual objects in the symbolic frames describe exactly
how the heap objects that have to be allocated on guard failure look like.
To this end, the content of the every field of the virtual object is described
in the same way that the local variables of symbolic frames are described.
The fields of the virtual objects can therefore be SSA variables, constants
or other virtual objects.

During the storing of resume data virtual objects are also shared
between subsequent guards as much as possible.
The same observation as about frames applies:
Quite often a virtual object does not change from one guard to the next.
Then the data structure is shared.

% subsection Interaction With Optimization (end)

@@

- frames consist of a PC and local variables

@@

As explained in previous sections, when a specific guard has failed often enogh
a new trace, refered to as bridge, starting from this guard is recorded and
compiled. When compiling bridges the goal is that future failures of the guards
that led to the compilation of the bridge should execute the bridge without
additional overhead, in particular it the failure of the guard should not lead
to leaving the compiled code prior to execution the code of the bridge.

The process of compiling a bridge is very similar to compiling a loop,
instructions and guards are processed in the same way as described above. The
main difference is the setup phase, when compiling a trace we start with a lean
slate. The compilation of a bridge is starts from a state (register and stack
bindings) that corresponds to the state during the compilation of the original
guard. To restore the state needed compile the bridge we use the encoded
representation created for the guard to rebuild the bindings from IR-variables
to stack locations and registers used in the register allocator.  With this
reconstruction all bindings are restored to the state as they were in the
original loop up to the guard.

Once the bridge has been compiled the trampoline method stub is redirected to
the code of the bridge. In future if the guard fails again it jumps to the code
compiled for the bridge instead of bailing out. Once the guard has been
compiled and attached to the loop the guard becomes just a point where
control-flow can split. The loop after the guard and the bridge are just
condional paths.
